Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-06-25 paper/manuscript.tex
additional citations

Because it is a discontinuous function of the position and orientation of the ligand relative to the receptor, molecular dynamics implementations often approximate the indicator function by means of flat-bottom restraints~\cite{chen2007can} (see below). Harmonic restraints are not as suitable in this respect because they favor a specific microstate at the bottom of the harmonic potential rather than specifying a region of conformational space.

@@

Binding site restraints implement the $I(\zeta)$ indicator function. They, therefore, can depend at most on the six rigid-body external coordinates of the ligand governing the position and orientation of the ligand relative to the receptor. Any other choice would affect the intramolecular conformational distribution of the ligand or the receptor and would  bias the binding free energy estimate. The external degrees of freedom introduced by Boresch et al.~\cite{boresch2003absolute,leitgeb2005alchemical} are a popular choice. The Boresch coordinates, as well as other similar approaches, involve selecting anchor atoms of the receptor and the ligand. In principle, the choice of anchor atoms is arbitrary, provided adequate simulation time is available, and as long as it leads to similar binding pose definitions. However, practical considerations may be important. For example, some choices of Boresch-style anchor atoms can be numerically unstable. Additionally, anchor atoms should likely be placed in a part of the molecule that defines the binding orientation well rather than in a floppy, solvent-exposed tail. Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite}.

@@

The Boresch harmonic restraints~\cite{boresch2003absolute,baumann2023broadening} are a popular choice as auxiliary restraints because they are based on interatomic distances, bond angles, and dihedral angles degrees of freedom already implemented in molecular mechanics codes that do not interfere with intramolecular conformational distributions. Moreover, the free energy of releasing Boresch restraints within the simulation box volume $V$ is easily computed analytically~\cite{boresch2003absolute}.

@@

Recently, degrees of freedom for binding site and auxiliary potentials based on larger sets of atomic positions of the receptor and ligand have been put forward.
Fu et al. suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand, which minimises the RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}. Salari et al. proposed restraining the ``distance-from-bound-configuration'' (DBC) variable, which is the RMSD of ligand coordinates in the frame of reference of the binding site~\cite{salari2018streamlined, ebrahimi2022symmetry}. DBC may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, making them unsuitable for describing the binding site macrostate and the corresponding standard state terms. However, these restraints are practical as auxiliary restraints where additional steps are used to introduce and release the DBC restraints. These restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing, clark2023comparison}.

>>>>

Because it is a discontinuous function of the position and orientation of the ligand relative to the receptor, molecular dynamics implementations often approximate the indicator function by means of flat-bottom restraints~\cite{chen2007can, wu2021alchemical} (see below). Harmonic restraints are not as suitable in this respect because they favor a specific microstate at the bottom of the harmonic potential rather than specifying a region of conformational space.

@@

Binding site restraints implement the $I(\zeta)$ indicator function. They, therefore, can depend at most on the six rigid-body external coordinates of the ligand governing the position and orientation of the ligand relative to the receptor. Any other choice would affect the intramolecular conformational distribution of the ligand or the receptor and would  bias the binding free energy estimate. The external degrees of freedom introduced by Boresch et al.~\cite{boresch2003absolute,leitgeb2005alchemical} are a popular choice. The Boresch coordinates, as well as other similar approaches, involve selecting anchor atoms of the receptor and the ligand. In principle, the choice of anchor atoms is arbitrary, provided adequate simulation time is available, and as long as it leads to similar binding pose definitions. However, practical considerations may be important. For example, some choices of Boresch-style anchor atoms can be numerically unstable. Additionally, anchor atoms should likely be placed in a part of the molecule that defines the binding orientation well rather than in a floppy, solvent-exposed tail. Several automated methods have been proposed to select anchor points for Boresch restraints, which typically involve analysing a simulation of the fully-interacting receptor-ligand complex~\cite{alibay2022evaluating, alibay2021mdrestraintsgenerator, baumann2023broadening, chen2023enhancing, hedges2023suite, wu2025optimizing}.

@@

The Boresch harmonic restraints~\cite{boresch2003absolute,baumann2023broadening} are a popular choice as auxiliary restraints because they are based on interatomic distances, bond angles, and dihedral angles degrees of freedom already implemented in molecular mechanics codes that do not interfere with intramolecular conformational distributions. Moreover, the free energy of releasing Boresch restraints within the simulation box volume $V$ is easily computed analytically~\cite{boresch2003absolute, boresch2024analytical}.

@@

Recently, degrees of freedom for binding site and auxiliary potentials based on larger sets of atomic positions of the receptor and ligand have been put forward.
Fu et al. suggested restraining the six relative rigid-body degrees of freedom derived by finding the rotation of the ligand, which minimises the RMSD to a reference structure of the receptor-ligand complex (after accounting for rotation and translation of the receptor)~\cite{fu2017new}. Salari et al. proposed restraining the ``distance-from-bound-configuration'' (DBC) variable, which is the RMSD of ligand coordinates in the frame of reference of the binding site~\cite{salari2018streamlined, ebrahimi2022symmetry}. DBC may allow the binding pose to be closely preserved as the ligand intermolecular interactions are removed, but has the disadvantage of coupling the internal and relative external degrees of freedom of the receptor and ligand, making them unsuitable for describing the binding site macrostate and the corresponding standard state terms. However, these restraints are practical as auxiliary restraints when additional steps are used to introduce and release them. Moreover, because they are based on flat-bottom potentials encompassing the desired binding poses, the free energy of releasing them in the bound state can be safely neglected in most cases. DBC restraint schemes are implemented in open-source workflows~\cite{fu2021bfee2, fu2022accurate, santiago-mcrae2023computing, clark2023comparison, hedges2023suite}.
